Regional Base-Flow Index in Arid Landscapes Using Machine Learning and Instrumented Records

### 2025-05-31 elsevier_MANUSCRIPT.qmd
Update elsevier_MANUSCRIPT.qmd

  - name: Benjamin Lucas
    orcid: 0000-0002-2021-3076
    corresponding: false
    roles: 
      - Conceptualization
      - Resources
      - Supervision
      - Software
      - Writing – Review & Editing 
    affiliations:
      - 'Department of Mathematics and Statistics, Northern Arizona University, Flagstaff, AZ, USA'
abstract: |
  Base flow, sustained by groundwater discharge, is a vital component of river ecosystems, particularly in drylands, where water resources are limited. This novel study analyzes the instrumented streamflow record in Arizona to assess long-term base-flow index (BFI) trends across gauged catchments. Results indicate that approximately 32% of Arizona’s streamflow originates from groundwater discharge, with significant spatial variability driven by landscape and climatic factors. Base-flow relationships are analyzed with coincident trends in climate variables such as precipitation, evapotranspiration, and temperature. Spatial and climatic trends reveal  variability in base-flow contributions, providing insight into groundwater-surface water interactions in arid and semi-arid landscapes. Building on this analysis, we applied machine learning methods to predict BFI in ungauged basins, addressing the challenges of Arizona’s sparse streamgage network. Using the eXtreme Gradient Boosting (XGBoost) algorithm trained and validated on observed hydroclimate and physiographic variables, we estimate long-term BFI from 1991 to 2020. This combined approach integrates observational data with predictive modeling to enhance our understanding of base-flow processes and provide a framework for water resource management in data-limited regions.

@@

Basin elevation was derived from a 30-meter resolution Digital Elevation Model (DEM) from the USGS National Elevation Dataset. The DEM of Arizona was used to derive key basin characteristics: basin area, average slope, and the proportion of each basin oriented toward north or south aspects. Various geospatial variables, such as aspect, were disaggregated then averaged to assess the areal percentage of each sub-variable within individual HUC 8 basins. By calculating these percentages, we derived a more comprehensive understanding of landscape composition across space. Land cover was acquired from USGS-NLCD (National Land Cover Database), hydrologic soil group from SSURGO (Soil Survey Geographic Database), and underlying geology and karst from USGS were all similarly averaged across the basins. Aggregating variables to align with the HUC 8 boundaries allowed for more precise predictions of BFI by integrating spatial variations within each basin.

@@

Machine learning (ML) approaches have demonstrated strong utility in hydrologic prediction tasks, including the estimation of streamflow and base-flow indices [@singh2018; @schmidt_ML_2020; @Rozos2021Machine]. In this study, we used the eXtreme Gradient Boosting (XGBoost) algorithm [@chen2016xgboost] to estimate annual BFI at ungauged locations across Arizona using catchment characteristics as predictors [@tbl-predictors]. XGBoost is a gradient boosting algorithm that constructs an ensemble of decision trees, where each tree iteratively corrects errors from previous trees to improve predictive accuracy [@chen2016xgboost]. The model was trained on annual BFI observations in order to estimate annual BFI in ungauged basins. These annual BFI estimates were averaged over time to represent long-term BFI, consistent with the time-integrated nature of our response metric.

@@

The training dataset consisted of 7,724 site-year observations, where each observation represents annual BFI for a single water year at a given USGS streamgage. These observations were paired with 45 predictor variables derived from basin characteristics and climate data. Predictor selection was based on prior hydrologic studies, data availability across the study area, and pairwise correlation screening to minimize multicollinearity. The predictors included basin-averaged climate metrics (e.g., precipitation, temperature), topographic features (e.g., elevation, slope), and land surface attributes (e.g., land cover classes, soil type, karst presence).

We tested multiple model configurations, including versions with the full predictor set across the state, separate models for physiographic regions with full predictor set, and reduced subsets across the state, to evaluate the impact of different predictor combinations on model performance and to ensure robustness of the final selected model. Dimensionality reduction was performed to improve interpretability and reduce computational burden. We applied SHAP (SHapley Additive exPlanations) values [@SHAP_values] to assess global feature importance and rank predictors. SHAP values quantify both the magnitude and direction of each feature’s contribution to predictions. The ten most important features—based on mean absolute SHAP value—were retained in a reduced model, which showed improved predictive accuracy and efficiency over the full model and regional models. Sensitivity analyses tested region-specific models and smaller predictor subsets; however, the state-wide model using SHAP-prioritized features consistently outperformed alternatives.

@@

The model dataset was evaluated using 10-fold cross-validation, ensuring robust performance estimates and guarding against overfitting [@fig-k-fold] . In each iteration (fold), the model was trained on 90% of the data and validated on the remaining 10%, with folds rotated until all data had been used for validation in separate models. Root mean squared error (RMSE) was used as the primary performance metric. Model performance on unseen data was strong, with an overall R² of 0.764. Additional model validation results are presented in section 3.2.1.

@@

Hyperparameter tuning was performed through an internal grid search. The search evaluated combinations of key parameters including the number of trees, learning rate ($\eta$), minimum split loss ($\gamma$), tree depth, and minimum child weight. A full discussion of these hyperparametes can be found in @chen2016xgboost. The final model used 700 trees, $\eta$ of 0.05, $\gamma$ of 0.075, a maximum depth of 7, and a minimum child weight of 5. All parameters were selected for their consistent reduction in RMSE across validation folds.

@@

#### Trends in BFI {#sec-bfi-trends}

Trends in long-term BFI over the period of record for each streamgage are illustrated in @fig-instrumented-trend and summarized in @tbl-trends. Long-term BFI trends were assessed at all instrumented sites using the Mann–Kendall test. As BFI quantifies the proportion of streamflow attributed to base flow, its long-term trends inherently reflect changes in subsurface contributions relative to total flow. A 72.2% coincidence rate between significant base flow and BFI trends supports the interpretation that changes in BFI are largely driven by underlying base-flow dynamics. Statistically significant long-term BFI trends were observed across numerous sites, with spatial variation highlighting regional differences in hydrologic response.

@@

Statistically significant decreasing trends in long-term BFI were more common than increasing trends across all site classifications (@tbl-trends). While decreasing trends dominate, both increasing and decreasing trends are observed within each classification. Monsoon-dominated regions exhibit a higher proportion of significant negative trends (24.1%) compared to snowmelt-dominated regions (10.2%), suggesting that monsoon-dominated systems are more consistently correlated with declining long-term BFI. Among climate classifications, warm-dry climates have the highest proportion of negative trends (20.0%), followed by warm-wet climates (19.4%), indicating that regions with higher temperatures are more prone to BFI declines. Low-slope regions show a greater prevalence of negative trends (20.4%) compared to high-slope regions (11.8%). This suggests that flatter areas may be more susceptible to base-flow reductions, potentially due to differences in hydrologic connectivity and recharge dynamics.

@@

Coincident trends are defined as those in which long-term BFI trends match in direction with climate variable trends over the same period. Results are summarized in @tbl-coincident-trends. This analysis includes both significant and non-significant trends, which is appropriate where the influence of complex, interconnected processes may not always manifest as statistically significant patterns over limited observational periods [@ficklin2016]. Coincidence was highest with precipitation (64.88% for base flow; 53.17% for BFI), followed by ETO and then temperature. Temperature trends most often show an inverse relationship—i.e., warming associated with declining BFI and base flow.

@@

The regionalized (HUC-8) long-term BFI (1991–2020) is shown in @fig-bfi-huc. Basins with high long-term BFIs, such as those along the Grand Canyon in the northwestern part of the study area, indicate greater surface water and groundwater interaction. Elevated BFI values are also observed along portions of the Mogollon Rim, a heavily forested region with high precipitation that marks the transition between physiographic regions. Additionally, headwater regions of perennial rivers tend to exhibit higher long-term BFI values. In contrast, low long-term BFI values are found in areas like the Defiance Plateau in northeastern Arizona and the arid southern regions of the state.

@@

Several previous studies have examined regional and continental patterns of BFI, offering useful context for our findings despite limited geographic overlap. @beck2013 developed a global map of BFI using an artificial neural network trained on 1,862 U.S. streamgages from the MOPEX dataset [@mopex_2006]. Only ten of these gauges overlap with our Arizona study area, limiting direct comparison. Nonetheless, the spatial distribution of BFI in Arizona from @beck2013 shows general agreement with our regional patterns. The higher BFI values reported by @beck2013 may reflect the influence of their global-scale ANN model and associated generalizations. Similarly, @ayers2022 examined base-flow dynamics using GAGES-II sites across CONUS, but their study included few gauges within Arizona. While exact locations are not specified, their published figures suggests regional agreement with our estimated BFI distributions. @santhi2008 employed interpolation methods for regional base-flow estimation across CONUS, but specific gauge locations or data are not available limiting comparison.\`\`\`\`

@@

This study addresses these monitoring gaps through an analysis of observed BFI at instrumented streamgages, highlighting groundwater-surface water interactions within Arizona’s arid and semi-arid landscapes. Observed spatial patterns in BFI clearly align with known hydrogeological and climatic variations. High long-term BFI values coincide with areas characterized by spring-fed streams and pronounced physiographic transitions, such as the Grand Canyon and Mogollon Rim. Conversely, lower observed BFI values correspond to regions with limited groundwater contributions, notably the Defiance Plateau and southern Arizona, emphasizing vulnerability in these arid environments. A trend of high BFI in upstream reaches to lower BFI in downstream reaches [@fig-instrumented-bfi] likely indicates a transition from zones of groundwater discharge to zones of groundwater recharge and may indicate a transition from gaining stream reaches to losing stream reaches [@winter2007].

Given the limited coverage provided by existing gauges, a common issue in arid and semi-arid regions globally [@krabbenhoft-2022], we utilized a machine learning approach to estimate annual BFI in ungauged catchments. The regional model exhibited strong predictive capability (overall R²=0.764), though systematic underprediction (negative pbias) was noted, particularly in monsoon-dominated and warm-dry climates. These biases suggest areas for improvement in model performance, potentially through the incorporation of additional predictors or more refined regional hydrologic representations. Although creating separate regional models (by physiographic region) was explored to enhance predictive accuracy, the state-wide model consistently outperformed these regionalized approaches, emphasizing the advantage of a comprehensive, state-wide dataset.

>>>>

  - name: Benjamin Lucas
    orcid: 0000-0002-2021-3076
    corresponding: false
    roles: 
      - Conceptualization
      - Resources
      - Supervision
      - Software
      - Writing – Review & Editing 
    affiliations:
      - 'Department of Mathematics and Statistics, Northern Arizona University, Flagstaff, AZ, USA'
abstract: |
  Base flow, sustained by groundwater discharge, is a vital component of river ecosystems, particularly in drylands, where water resources are limited. This novel study analyzes the instrumented streamflow record in Arizona to assess long-term base-flow index (BFI) trends across gauged catchments. Results indicate that approximately 32% of Arizona’s streamflow originates from groundwater discharge, with significant spatial variability driven by landscape and climatic factors. Base-flow relationships are analyzed with coincident trends in climate variables such as precipitation, evapotranspiration, and temperature. Spatial and climatic trends reveal  variability in base-flow contributions, providing insight into groundwater-surface water interactions in arid and semi-arid landscapes. Building on this analysis, we applied machine learning methods to predict BFI in ungauged basins, addressing the challenges of Arizona’s sparse streamgage network. Using the eXtreme Gradient Boosting (XGBoost) algorithm trained and validated on observed hydroclimate and physiographic variables, we estimate long-term BFI from 1991 to 2020. Results indicate spatial heterogeneity in BFI trends, with decreasing baseflow most pronounced in warm-dry and monsoon-dominated catchments. This combined approach integrates observational data with predictive modeling to enhance our understanding of base-flow processes and provide a framework for water resource management in data-limited regions.

@@

Basin elevation was derived from a 30-meter resolution Digital Elevation Model (DEM) from the USGS National Elevation Dataset (NED). The DEM of Arizona was used to derive key basin characteristics: basin area, basin slope, and the proportion of each basin oriented toward north or south aspects. Various geospatial variables, such as aspect, were disaggregated then averaged to assess the areal percentage of each sub-variable within individual HUC 8 basins. By calculating these percentages, we derived a more comprehensive understanding of landscape composition across space. Land cover was acquired from USGS-NLCD (National Land Cover Database), hydrologic soil group from SSURGO (Soil Survey Geographic Database), and underlying geology and karst from USGS were all similarly averaged across the basins. Aggregating variables to align with the HUC 8 boundaries allowed for more precise predictions of BFI by integrating spatial variations within each basin.

@@

Machine learning (ML) approaches have demonstrated strong utility in hydrologic prediction tasks, including the estimation of streamflow and base-flow indices [@singh2018; @schmidt_ML_2020; @Rozos2021Machine]. In this study, we used the eXtreme Gradient Boosting (XGBoost) algorithm [@chen2016xgboost] to estimate annual BFI at ungauged locations across Arizona using catchment characteristics as predictors [@tbl-predictors]. XGBoost is a gradient boosting algorithm that builds an ensemble of decision trees, where each tree iteratively corrects errors from previous trees to improve predictive accuracy [@chen2016xgboost]. The model was trained on annual BFI observations in order to estimate annual BFI in ungauged basins. These annual BFI estimates were averaged over time to represent long-term BFI, consistent with the time-integrated nature of our response metric.

@@

The training dataset consisted of 7,724 site-year observations, where each observation represents annual BFI for a single year at a given USGS streamgage. These observations are paired with 45 predictor variables derived from basin characteristics and climate data. Predictor selection was based on prior hydrologic studies and data availability across the study area. The predictors included basin-averaged climate metrics (e.g., precipitation, temperature), topographic features (e.g., elevation, slope), and land surface attributes (e.g., land cover classes, soil type, presence of karst).

We tested multiple model configurations, including versions with the full predictor set across the state, separate models for physiographic regions with full predictor set, and reduced subsets across the state, to evaluate the impact of different predictor combinations on model performance and to ensure robustness of the final selected model. Dimensionality reduction was performed to improve interpretability and reduce computational burden. We applied SHAP (SHapley Additive exPlanations) values [@SHAP_values] to assess global feature importance and rank predictors. SHAP values quantify both the magnitude and direction of each feature’s contribution to predictions. The ten most important features, based on SHAP value, were retained in a reduced model. The reduced model showed improved predictive accuracy and efficiency over the full model and regional models. Sensitivity analyses tested region-specific models and smaller predictor subsets; however, the state-wide model using SHAP-prioritized features outperformed alternatives. The final model was applied to all HUC8 basins across Arizona using the same predictor layers to estimate long-term BFI values.

@@

The model dataset was evaluated using 10-fold cross-validation, ensuring robust performance estimates and guarding against overfitting [@fig-k-fold] . In each iteration (fold), the model was trained on 90% of the data and validated on the remaining 10%, with folds being rotated until all of the data had been used for validation in separate models. Root mean squared error (RMSE) was used as the primary performance metric. Model performance on unseen data was strong, with an overall R² of 0.764. Additional model validation results are presented in section 3.2.1.

@@

Hyperparameter tuning was performed through an internal grid search. The search evaluated combinations of parameters including the number of trees, learning rate ($\eta$), minimum split loss ($\gamma$), tree depth, and minimum child weight. A description of these hyperparametes can be found in @chen2016xgboost. The final model used 700 trees, $\eta$ of 0.05, $\gamma$ of 0.075, a maximum depth of 7, and a minimum child weight of 5. All parameters were selected for their reduction in RMSE across validation folds.

@@

#### Trends in Annual BFI {#sec-bfi-trends}

Trends in annual BFI over the period of record for each streamgage are illustrated in @fig-instrumented-trend and summarized in @tbl-trends. Annual BFI trends were assessed at all instrumented sites using the Mann–Kendall test. As BFI quantifies the proportion of streamflow attributed to base flow, its long-term trends inherently reflect changes in subsurface contributions relative to total flow. A 72.2% coincidence rate between significant base-flow volume and BFI trends supports the interpretation that changes in BFI are largely driven by underlying base-flow dynamics. Statistically significant long-term BFI trends were observed across numerous sites, with spatial variation highlighting regional differences in hydrologic response.

@@

Statistically significant decreasing trends in long-term BFI were more common than increasing trends across all site classifications (@tbl-trends). While decreasing trends dominate, both increasing and decreasing trends are observed within each classification. Monsoon-dominated regions exhibit a higher proportion of significant negative trends (24.1%) compared to snowmelt-dominated regions (10.2%), suggesting that monsoon-dominated systems are more consistently correlated with declining long-term BFI. Among climate classifications, warm-dry climates have the highest proportion of negative trends (20.0%), followed by warm-wet climates (19.4%), indicating that regions with higher temperatures are more prone to BFI declines. Low-slope regions show a greater prevalence of negative trends (20.4%) compared to high-slope regions (11.8%). This suggests that flatter areas may be more susceptible to base-flow reductions, potentially due to differences in hydrologic connectivity and recharge dynamics. Negative trends in low-slope regions may also be a result of reductions in water table height, assumed to follow topography, leading to a change in flow direction.

@@

Coincident trends are defined as those in which long-term BFI trends match in direction with climate variable trends over the same period. Results are summarized in @tbl-coincident-trends. This analysis includes both significant and non-significant trends, which is appropriate where the influence of complex, interconnected processes may not always manifest as statistically significant patterns over limited observational periods [@ficklin2016]. Long-term BFI coincided most with precipitation (53.17%), followed by ET~O~ and then temperature. Temperature trends most often show an inverse relationship—i.e., warming associated with declining BFI and base flow.

@@

Predicted long-term BFI (1991–2020) across Arizona's HUC-8 basins is shown in @fig-bfi-huc. Higher values were observed along the Grand Canyon corridor and parts of the Mogollon Rim, while lower values were found across the Defiance Plateau and in southern Arizona. Spatial variation generally aligns with physiographic regions and precipitation gradients, with upstream reaches of perennial rivers typically exhibiting higher BFI.

@@

Given the limited coverage provided by existing gauges, a common issue in arid and semi-arid regions globally [@krabbenhoft-2022], we utilized a machine learning approach to estimate annual BFI in ungauged catchments. The regional model exhibited strong predictive capability (overall R²=0.764), though systematic underprediction (negative pbias) was noted, particularly in monsoon-dominated and warm-dry climates. These biases highlight opportunities to improve model performance through incorporating additional predictors or refining regional hydrologic representations. Although creating separate regional models (by physiographic region) was explored to enhance predictive accuracy, the state-wide model consistently outperformed these regionalized approaches, emphasizing the advantage of a comprehensive, state-wide dataset.

The spatial patterns in BFI reflect underlying differences in hydrogeologic and climatic drivers. In high-BFI regions such as the Mogollon Rim and Grand Canyon, fractured bedrock, karst systems, and deep regional aquifers provide substantial groundwater discharge to surface waters. These areas benefit from high orographic precipitation and forested landscapes that promote infiltration and recharge. In contrast, low-BFI regions such as the Defiance Plateau and southern Arizona feature shallow or disconnected groundwater systems, low precipitation, and high evaporative demand, all of which limit baseflow contributions. Land cover and soil properties further modulate these processes; areas with high-infiltration soils and perennial vegetation are more conducive to sustained baseflow, while impervious surfaces and fine-grained soils suppress it. The downstream decline in BFI [@fig-instrumented-bfi] is consistent with the transition from headwater recharge zones to more arid, losing stream reaches, highlighting the importance of hydrogeologic setting in shaping baseflow dynamics [@winter2007].

Several previous studies have examined regional and continental patterns of BFI, offering useful context for our findings despite limited geographic overlap. @beck2013 developed a global map of BFI using an artificial neural network trained on 1,862 U.S. streamgages from the MOPEX dataset [@mopex_2006]. Only ten of these gauges overlap with our Arizona study area, limiting direct comparison. Nonetheless, the spatial distribution of BFI in Arizona from @beck2013 shows general agreement with our regional patterns. The higher BFI values reported by @beck2013 may reflect the influence of their global-scale ANN model and associated generalizations. Similarly, @ayers2022 examined base-flow dynamics using GAGES-II sites across CONUS, but their study included few gauges within Arizona. While exact locations are not specified, their published figures suggests regional agreement with our estimated BFI distributions. @santhi2008 employed interpolation methods for regional base-flow estimation across CONUS; however, the lack of gauge-level detail limits direct comparison.

@@
